\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-12 analysis/paper/paper.tex

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
main author's website, and it is also available as part of the
\pkg{casebase} package.

@@

\begin{CodeChunk}
\begin{table}[t]

\caption{\label{tab:table1bmtcrr}Baseline characteristics of patients in the stem-cell transplant study.}
\centering
\begin{tabular}{lll}
\toprule
Variable & Description & Statistical summary\\
\midrule
\code{Sex} & Sex & M=Male (100)\\
 &  & F=Female (77)\\
\code{D} & Disease & ALL (73)\\
 &  & AML (104)\\
\code{Phase} & Phase & CR1 (47)\\
\addlinespace
 &  & CR2 (45)\\
 &  & CR3 (12)\\
 &  & Relapse (73)\\
\code{Source} & Type of transplant & BM+PB (21)\\
 &  & PB (156)\\
\addlinespace
\code{Age} & Age of patient (years) & 4--62\\
 &  & 30.47 (13.04)\\
\code{ftime} & Failure time (months) & 0.13--131.77\\
 &  & 20.28 (30.78)\\
\code{Status} & Status indicator & 0=censored (46)\\
\addlinespace
 &  & 1=relapse (56)\\
 &  & 2=competing event (75)\\
\bottomrule
\end{tabular}
\end{table}

@@

{\centering \includegraphics{../figures/compPop-1}

\caption[Population-time plot for the stem-cell transplant study]{Population-time plot for the stem-cell transplant study. A) The red points represent relapse events. B) The blue points represent the competing events.}\label{fig:compPop}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R+                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{table}[t]

\caption{\label{tab:bmtcrr-cis}ADD CAPTION}
\centering
\begin{tabular}{lrlrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Case-Base} & \multicolumn{2}{c}{Cox} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Covariates & HR & 95\% CI & HR & 95\% CI\\
\midrule
Sex & 0.77 & (0.45, 1.35) & 0.68 & (0.39, 1.19)\\
Disease & 0.58 & (0.32, 1.04) & 0.52 & (0.29, 0.93)\\
Phase (CR2 vs. CR1) & 1.11 & (0.45, 2.76) & 1.21 & (0.48, 3.02)\\
Phase (CR3 vs. CR1) & 1.58 & (0.41, 6.04) & 1.67 & (0.43, 6.5)\\
Phase (Relapse vs. CR1) & 3.89 & (1.81, 8.35) & 4.55 & (2.09, 9.9)\\
\addlinespace
Source & 1.48 & (0.49, 4.46) & 1.46 & (0.47, 4.54)\\
Age & 0.99 & (0.97, 1.02) & 0.99 & (0.97, 1.02)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R+                                      levels = levels(bmtcrr[,"Sex"])),
R+                       "D" = c("ALL", "AML"), # Both diseases
R+                       "Phase" = factor(c("Relapse", "Relapse"), 
R+                                        levels = levels(bmtcrr[,"Phase"])),
R+                       "Age" = c(35, 35),
R+                       "Source" = factor(c("PB", "PB"), 
R+                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R+                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R+                           const(Phase) + const(Source) + const(Age), 
R+                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

{\centering \includegraphics{../figures/bmtcrr-risk-1}

\caption{\label{fig:compAbsrisk}Absolute risk curve for a fixed covariate profile and the two disease groups. The estimate obtained from case-base sampling is compared to the Kaplan-Meier estimate.}\label{fig:bmtcrr-risk}
\end{figure}
\end{CodeChunk}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using
predictive mean matching and using the \pkg{mice} package in
\proglang{R}. Before imputation, there was a total of 1000 individuals
and 35 variables. After imputation, we have 474 observations where
67.5\% of individuals in the study experience the event of interest.
\textcolor{red}{How come we lose individuals during imputation?} The
SUPPORT dataset will be used to demonstrate regularization within
casebase, while comparing their absolute risk predictions for a new
individual. A description of each variable can be found in Table
\ref{tab:support1} and a breakdown of each categorical variable in Table
\ref{tab:support2}. The data was downloaded from the main author's
website, and it is also available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package on CRAN was used to introduce
elastic net, lasso and ridge regressions to casebase. To visualize the
incidence density, we can refer to the post-imputation population-time
plot in Figure \ref{fig:ptPostI}.

@@

\begin{CodeInput}
R> popTime_pre <- popTime(as.data.frame(support),
R>                        time = "d.time", event = "death")
R> plot(popTime_pre)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> popTime_post <- popTime(as.data.frame(completeData),
R>                         time = "d.time", event = "death")
R> plot(popTime_post)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given th number of covariates available, we
will do this using regularized estimation. First, we fit the hazard
function; for the sake of simplicity, we opted for a linear term for
time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will
use the function \code{fitsmoothhazard.fit}, which follows the syntax
from \code{glmnet}. Specifically, we input a matrix \code{y}, which
contains the time and event variables, and a matrix \code{x}, which
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting \code{alpha} to
1.

@@

\begin{CodeInput}
R> cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
R>                                 event = "death", formula_time = ~ d.time, 
R>                                 alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function.

@@

\begin{CodeInput}
R> 
R> casebaseAbsolute <- absoluteRisk(cb.Model,
R>                                  time = seq(1, max(completeData$d.time), 0.5),
R>                                  newdata = t(newData),
R>                                  s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation. However,
\pkg{glmnet} does not provide any functionality for estimating the
baseline hazard, and the functionalities in the \pkg{survival} package
do not work on the output from \code{cv.glmnet}. To circumvent this
issue and estimate the baseline hazard, we followed the following steps.
First, we fitted a non-regularized Cox model using only the variables
selected by \pkg{glmnet}; the baseline hazard will be estimated using
this object. Then, we replaced the estimates of the regression
coefficients by the estimates from \pkg{glmnet}. Finally, this modified
\code{coxph} object is fed to the \code{survival::survfit} function to
obtain the estimated survival function.

@@

\begin{CodeInput}
R> # Create u and xCox, which will be used when fitting Cox with glmnet.
R> u <- survival::Surv(time = as.numeric(y[,2]), 
R>                     event = as.factor(y[,1]))
R> xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
R>                                       data = completeData))[-c(sam),]
R> # hazard for cox using glmnet
R> coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
R>                                family = "cox", alpha = 1)
R> # convergence demonstrated in plot
R> # plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> # creating a new dataset that only contains the covariates chosen through glmnet.
R> cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
R>                                     as.numeric(y[,1]),
R>                                     xCox[, nonzero_covariate_cox$X1]))
R> # fitting a cox model using regular estimation, however we will not keep it.
R> # this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
R>                           data = cleanCoxData)
R> # The coefficients of this object will be replaced 
R> # with the estimates from coxglmFit.
R> coxFit$coefficients <- nonzero_coef_cox@x
R> # Fitting CI curve for cox+glmnet
R> newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
R> 
R> abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
R>                               time = seq(0, max(completeData$d.time), 1),
R>                               type = "breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> km <- survival::Surv(time = completeData$d.time, 
R>                      event = completeData$death)
R> abKm <- survival::survfit(km ~ 1,
R>                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.

\begin{CodeInput}
R> plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
R>      main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
R>      xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
R> lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
R> lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
R> legend("bottom", 
R>        legend = c("casebase", "Cox+glmnet", "KM curve"), 
R>        col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

This report was generated on 2019-12-12 13:11:14 using the following
computational environment and dependencies:

@@

#>  cowplot     * 1.0.0      2019-07-11 [1]

@@

#>  glue        * 1.3.1      2019-03-12 [1]

@@

#>  lava          1.6.5      2019-02-12 [1]

@@

#>  numDeriv      2016.8-1   2016-08-27 [1]

@@

#>  plyr          1.8.5      2019-12-10 [1]

@@

#>  reshape2      1.4.3      2017-12-11 [1]

@@

#>  timereg     * 1.9.4      2019-07-30 [1]

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [c4f42a9] 2019-12-11: Merge branch 'review-glmnet' into review
\end{CodeOutput}
\end{CodeChunk}

>>>>

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

@@

@@

@@

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"), # Both diseases
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R>                           const(Phase) + const(Source) + const(Age), 
R>                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

@@

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
more advanced features.

For the third case study, we will look at regularized estimation for the
hazard function. This can be performed using a combination of case-base
sampling and regularized logistic regression. To demonstrate this
capability, we examine the Study to Understand Prognoses Preferences
Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using predictive mean matching and using the
\pkg{mice} package in \proglang{R} \citep{mice}. Before imputation,
there was a total of 1000 individuals and 35 variables. After
imputation, we have 474 individuals.
\textcolor{red}{How come we lose individuals during imputation?} Of
these, 67.5\% experienced the event of interest. A description of each
variable can be found in Table \ref{tab:support1} and a breakdown of
each categorical variable in Table \ref{tab:support2}. The original data
was obtained from the website of the Department of Biostatistics at
Vanderbilt University. The imputed dataset is available as part of the
\pkg{casebase} package.

@@

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.
\textcolor{red}{This population-time plot is strange. The red dots never seem to go above 200.}

@@

\begin{CodeInput}
R> data(support)
R> support_popTime <- popTime(support,
R+                            time = "d.time", event = "death")
R> plot(support_popTime)
\end{CodeInput}
\begin{figure}

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for tje SUPPORT data.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before we proceed with the analysis, we first remove hospital death as a
covariate, as this is directly informative of death and cannot be used
as a predictor. Then we select one observation at random. We will
estimate the cumulative incidence curve for this individual. The other
observations will be used to fit the hazard function.

@@

\begin{CodeInput}
R> support <- dplyr::select(support, -hospdead)
R> 
R> # Randomly select one observation
R> indiv <- sample(1:nrow(support), size = 1)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given the number of covariates available, we
will do this using regularized estimation. As \pkg{casebase} makes use
of the \pkg{glmnet} package, we will use the function
\code{fitsmoothhazard.fit}, which follows the syntax from \code{glmnet}.
Specifically, we input a matrix \code{y}, which contains the time and
event variables, and a matrix \code{x}, which contains all other
variables we would like to include in the model. For the SUPPORT
dataset, all factor variables were converted into dummy variables.

@@

\begin{CodeInput}
R> # Create design matrix without the time variable
R> support_matrix <- model.matrix(death ~ . - d.time, 
R+                                data = support)
R> # Remove intercept term
R> support_matrix <- support_matrix[,colnames(support_matrix) != "(Intercept)"]
R> 
R> # Create test and train dataset
R> support_test <- support_matrix[indiv,,drop = FALSE]
R> 
R> support_train <- support_matrix[-indiv,]
R> y <- as.matrix(support[-indiv, c("death", "d.time")])
\end{CodeInput}
\end{CodeChunk}

We are now ready to estimate the hazard function. We use lasso
penalization; this is achieved by setting \code{alpha} to 1. Also, note
that the argument \code{formula_time} can be used to specify the
function form for time; in our example, we model time linearly.

@@

\begin{CodeInput}
R> support_cb <- fitSmoothHazard.fit(support_train, y, 
R+                                   formula_time = ~ d.time, 
R+                                   family = "glmnet", 
R+                                   event = "death", time = "d.time", 
R+                                   alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function. Since
\code{glmnet} fitted the penalized logistic regression for a range of
penalty terms, we need to specify which one we want to use for our
prediction. We select \code{s = "lambda.1se"}, as recommended.

@@

\begin{CodeInput}
R> time_points <- seq(1, max(support$d.time), by = 0.5)
R> support_risk <- absoluteRisk(support_cb,
R+                              time = time_points,
R+                              newdata = support_test,
R+                              s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation.

\begin{CodeInput}
R> # Create survival object for glmnet
R> surv_glmnet <- survival::Surv(time = y[, "d.time"], 
R+                               event = y[, "death"])
R> 
R> support_cox <- glmnet::cv.glmnet(x = support_train, 
R+                                  y = surv_glmnet,
R+                                  family = "cox", alpha = 1)
\end{CodeInput}
\end{CodeChunk}

However, \pkg{glmnet} does not provide any functionality for estimating
the baseline hazard, and the functionalities in the \pkg{survival}
package do not work on the output from \code{cv.glmnet}. To circumvent
this issue and get an estimate of the baseline hazard, we followed the
following steps. First, we fitted a non-regularized Cox model using only
the variables selected by \pkg{glmnet}; the baseline hazard will be
estimated using this object. Then, we replaced the estimates of the
regression coefficients by the estimates from \pkg{glmnet}. Finally,
this modified \code{coxph} object is fed to the \code{survival::survfit}
function to obtain the estimated survival function.

\begin{CodeInput}
R> # Identifying selected covariates
R> nonzero_covariates_cox <- predict(support_cox, type = "nonzero", 
R+                                   s = "lambda.1se")
R> nonzero_coef_cox <- coef(support_cox, s = "lambda.1se")
R> # Creating a new dataset that only contains the covariates selected by glmnet
R> support_cox_data <- as.data.frame(cbind(y,
R+                                         support_train[, nonzero_covariates_cox$X1]))
R> # Fitting a cox model
R> support_cox <- survival::coxph(Surv(time = d.time, event = death) ~ ., 
R+                           data = support_cox_data)
R> # Replace the estimate with those obtained by glmnet
R> support_cox$coefficients <- nonzero_coef_cox@x
R> 
R> # Fitting CI curve for cox+glmnet
R> support_cox_risk <- survival::survfit(support_cox,
R+                                       newdata = as.data.frame(support_test),
R+                                       time = time_points,
R+                                       type = "breslow")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> km <- survival::Surv(time = support$d.time, 
R+                      event = support$death)
R> abKm <- survival::survfit(km ~ 1,
R+                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.
\textcolor{red}{It would be good to have an idea of why the case-base and Coxnet curves are so different. At the very least, I think we should compare the select covariates.}

{\centering \includegraphics{../figures/abSupportComparison-1}

\caption{\label{fig:abSC} Comparison of cumulative incidence curves between regularized casebase in black, regularized Cox regression in red and Kaplan-Meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

@@

This report was generated on 2019-12-12 15:16:14 using the following
computational environment and dependencies:

@@

@@

#>  glue          1.3.1      2019-03-12 [1]

@@

@@

@@

@@

@@

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [c095709] 2019-12-12: review 2nd case study
\end{CodeOutput}
\end{CodeChunk}
